---
title: Investigating variance estimation via the bootstrap for simulated double observer data
author: "Matthew Murchie"
date: "17 June 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Variance estimation of abundance is a key component in distance sampling. The uncertainty associated with abundance estimates is affected by two main components:
   
1. *Detection function*, variation due to estimating the parameters of the      detection function.
2. *Encounter rate*, variation in the number of animals observed if we were to repeat our survey.
   
The ```dht``` function, found in the package ```mrds```, calculates the first component listed above using the delta method and the second component using one of three methods. See ```?dht``` for further details.   
In order to analyse the performance of the bootstrap it is necessary to simulate double observer data, where the true abundance and detection function is known.

## Simulating double observer data [^1]

### Region 
We begin by creating a region object. For simplicity, we will consider a square survey region with a total area of 1km2.
Using the ```DSsim``` (Marshall 2014) library in R, you can specify the coordinates of the survey region and any areas to be excluded.

```{r region, message=FALSE}
library(DSsim)
library(knitr)

coords <- gaps <- list()
coords[[1]] <- list(data.frame(x = c(0,1000,1000,0,0), y = c(0,0,1000,1000,0)))
gaps[[1]] <- list(data.frame(x = c(0,0,0,0,0), y = c(0,0,0,0,0)))

# create region object
region <- make.region(region.name = "study.area", units = "m",
                      coords = coords, gaps = gaps)
```

### Density 
Next we create a density object by specifying a grid with a constant density, adding hotspots accordingly.
```{r density}
pop.density <- make.density(region.obj = region, x.space = 10,
                            y.space = 10, constant = 0.5)
pop.density<- add.hotspot(pop.density, centre = c(500, 500),
                          sigma = 1000, amplitude = 0.2)
```

### Survey Design 
The transects are first generated as shapefiles using the ```write.transects``` function [^2], which are then saved to a specified directory.     
Note that you will need to save ```write.transects``` as a file and use ```source``` to read the function into R. 

``` {r write.transects, include=FALSE}
source("C:/Users/Matthew/Documents/SUMMER SCHOOL/MRDS_bootstrap_variance_estimation/R/write.transects.R")
```

``` {r write.transects eg, eval=FALSE}
# specify write.transects path directory
source("C:/...")
```

Here we generate three horizontal transects that span the length of the survey region. By changing the x and y coordinates and the number of sequence replicates, you can specify the number of transects together with their location.

``` {r transects}
n_segs <- 2
lines.many <- data.frame(x   = c(rep(seq(0, 1000, len=n_segs), 4)),
                         y   = c(seq(200, 200, len=n_segs),
                                 seq(400, 400, len=n_segs),
                                 seq(600, 600, len=n_segs),
                                 seq(800, 800, len=n_segs)),
                         leg = c(rep("1", n_segs),
                                 rep("2", n_segs),
                                 rep("3", n_segs),
                                 rep("4", n_segs)))
```

Before the survey shapefiles can be generated, you will need to create an empty folder called *shapes* where the shapefiles can be saved. The object ```shapefile.pathway``` specifies the location of this folder.   
Ignore the warning message from ```write.transects```. Here I have used the ```unlink``` function to ensure the *shapes* folder is empty. To check the shapefiles have been generated simply open the *shapes* folder, where you should find three new files.

``` {r shapefile directory, include=FALSE}
shapefile.pathway <- "C:/Users/Matthew/Documents/SUMMER SCHOOL/MRDS_bootstrap_variance_estimation/shapes"
```

``` {r shapefile eg, eval=FALSE}
# specify path to shapes folder
shapefile.pathway <- "C:/.../shapes"
```

``` {r write transects, warning=FALSE}
unlink(shapefile.pathway, recursive = TRUE)
write.transects(lines.many, shapefile.pathway)
```

Having generated and saved the shapefiles, we supply the ```make.design``` function with the path to the *shapes* folder to create the survey design object. 

``` {r survey design}
survey.design <- make.design(transect.type = "Line",
                             design.details = c("Parallel","Systematic"), 
                             region = region, plus.sampling = FALSE, 
                             path = shapefile.pathway)
```

### Abundance
Here we will fix the true abundance of animals to be 500. It is important to specify the argument ```fixed.N = TRUE```, when using the ```make.population.description``` function, to ensure this exact population size is generated with each realisation.

``` {r abundance}
pop.description <- make.population.description(N = 500, density.obj = pop.density,
                                               region = region, fixed.N = TRUE)
```

### Analyses 

The analyses object describes the model to be fitted to the distance data. Below we propose half-normal and hazard-rate models, using AIC to choose between them.

``` {r analyses}
ddf.analyses <- make.ddf.analysis.list(dsmodel = list(~cds(key = "hn",formula = ~1),
                                                      ~cds(key = "hr", formula = ~1)),
                                       method = "ds", criteria = "AIC", truncation = 30)
```

### Detection function

As we are simulating double observer data, we need to specify different detection functions for both observers. In both cases we select a half-normal detection function, adjusting the scale parameters associated with each.

``` {r det function}
detect.1 <- make.detectability(key.function = "hn", scale.param = 20,
                               truncation = 30)

detect.2 <- make.detectability(key.function = "hn", scale.param = 5,
                               truncation = 30)
```

### Simulation

Now we are in a position to simulate double observer data. Using the ```double.obs.sim``` function [^3], we pull together all the survey objects created above.
If ```plot=TRUE```, you can simulate the detection process for each observer.

``` {r double.obs.sim, include=FALSE}
source("C:/Users/Matthew/Documents/SUMMER SCHOOL/MRDS_bootstrap_variance_estimation/R/double.obs.sim.R")
```

``` {r double.obs.sim eg, eval=FALSE}
# specify write.transects path directory
source("C:/...")
```

``` {r simulation, fig.width=10, fig.height=4}
# two plots in window
par(mfrow=c(1,2))

tables <- double.obs.sim(region, survey.design, pop.description, detect.1,
                         detect.2, ddf.analyses, plot=TRUE)
```

``` {r window options, include=FALSE}
par(mfrow=c(1,1))
```

## Variance estimation via the bootstrap
To investigate variance estimation via the bootstrap, compared to methods currently implemented in ```dht```, we will consider three factors:   

1. the number of transects used in the survey design, 
2. the number of animals in the survey region (abundance) and 
3. the distribution/density of animals.  

The survey region and design will be the same as simulated above, with changes being made to the factor in question.

### Survey design
We will consider survey designs of 3, 9 and 15 transects in total. For brevity, I omit the code used to generate the transects - the approach is the same as above.   
The plots given below represent each survey design.

```{r transect plots, echo=FALSE, fig.width=10, fig.height=4}
par(mfrow=c(1,3))

plot(region)
abline(h=c(250,500,750), col = 4, lwd = 2)

plot(region)
abline(h=seq(100,900,100), col = 4, lwd = 2)

plot(region)
abline(h=seq(62.5,937.5,62.5), col = 4, lwd = 2)
```

Below are the results for the survey design with only three transects.

``` {r three transects, echo=FALSE, warning=FALSE}
n_segs <- 2

source("C:/Users/Matthew/Documents/SUMMER SCHOOL/MRDS_bootstrap_variance_estimation/R/bootstrap function.R")

detect.1 <- make.detectability(key.function = "hn", scale.param = 50,
                               truncation = 30)

detect.2 <- make.detectability(key.function = "hn", scale.param = 40,
                               truncation = 30)

# three transects
lines.low <- data.frame(x   = c(rep(seq(0, 1000, len=n_segs), 3)),
                        y   = c(seq(250, 250, len=n_segs),
                                seq(500, 500, len=n_segs),
                                seq(750, 750, len=n_segs)),
                        leg = c(rep("1", n_segs),
                                rep("2", n_segs),
                                rep("3", n_segs)))

unlink(shapefile.pathway, recursive = TRUE)
write.transects(lines.low, shapefile.pathway)

survey.design <- make.design(transect.type = "Line",
                             design.details = c("Parallel","Systematic"), 
                             region = region, plus.sampling = FALSE, 
                             path = shapefile.pathway)

tables <- double.obs.sim(region, survey.design, pop.description, detect.1,
                         detect.2, ddf.analyses, seed = 999)

tables$data["Region.Label"] <- rep("study.area", length(tables$data$object))


results <- boot.dht(tables, B=49, trunc=30)
results.df <- rbind(results$original, results$bootstrap)
kable(results.df)

```

Below are the results for the survey design with nine transects.

``` {r nine transects, echo=FALSE, warning=FALSE}
# nine transects
lines.med <- data.frame(x   = c(rep(seq(0, 1000, len=n_segs), 9)),
                        y   = c(seq(100, 100, len=n_segs),
                                seq(200, 200, len=n_segs),
                                seq(300, 300, len=n_segs),
                                seq(400, 400, len=n_segs),
                                seq(500, 500, len=n_segs),
                                seq(600, 600, len=n_segs),
                                seq(700, 700, len=n_segs),
                                seq(800, 800, len=n_segs),
                                seq(900, 900, len=n_segs)),
                        leg = c(rep("1", n_segs),
                                rep("2", n_segs),
                                rep("3", n_segs),
                                rep("4", n_segs),
                                rep("5", n_segs),
                                rep("6", n_segs),
                                rep("7", n_segs),
                                rep("8", n_segs),
                                rep("9", n_segs)))

unlink(shapefile.pathway, recursive = TRUE)
write.transects(lines.med, shapefile.pathway)

survey.design <- make.design(transect.type = "Line",
                             design.details = c("Parallel","Systematic"), 
                             region = region, plus.sampling = FALSE, 
                             path = shapefile.pathway)

tables <- double.obs.sim(region, survey.design, pop.description, detect.1,
                         detect.2, ddf.analyses, seed = 999)

tables$data["Region.Label"] <- rep("study.area", length(tables$data$object))


results <- boot.dht(tables, B=49, trunc=30)
results.df <- rbind(results$original, results$bootstrap)
kable(results.df)

```

Below are the results for the survey design with fifteen transects.

``` {r fifteen transects, echo=FALSE, warning=FALSE}
# fifteen transects
lines.high <- data.frame(x   = c(rep(seq(0, 1000, len=n_segs), 15)),
                         y   = c(seq(62.5, 62.5, len=n_segs),
                                 seq(125, 125, len=n_segs),
                                 seq(187.5, 187.5, len=n_segs),
                                 seq(250, 250, len=n_segs),
                                 seq(312.5, 312.5, len=n_segs),
                                 seq(375, 375, len=n_segs),
                                 seq(437.5, 437.5, len=n_segs),
                                 seq(500, 500, len=n_segs),
                                 seq(562.5, 562.5, len=n_segs),
                                 seq(625, 625, len=n_segs),
                                 seq(687.5, 687.5, len=n_segs),
                                 seq(750, 750, len=n_segs),
                                 seq(812.5, 812.5, len=n_segs),
                                 seq(875, 875, len=n_segs),
                                 seq(937.5, 937.5, len=n_segs)),
                         leg = c(rep("1", n_segs),
                                 rep("2", n_segs),
                                 rep("3", n_segs),
                                 rep("4", n_segs),
                                 rep("5", n_segs),
                                 rep("6", n_segs),
                                 rep("7", n_segs),
                                 rep("8", n_segs),
                                 rep("9", n_segs),
                                 rep("10", n_segs),
                                 rep("11", n_segs),
                                 rep("12", n_segs),
                                 rep("13", n_segs),
                                 rep("14", n_segs),
                                 rep("15", n_segs)))

unlink(shapefile.pathway, recursive = TRUE)
write.transects(lines.high, shapefile.pathway)

survey.design <- make.design(transect.type = "Line",
                             design.details = c("Parallel","Systematic"), 
                             region = region, plus.sampling = FALSE, 
                             path = shapefile.pathway)

tables <- double.obs.sim(region, survey.design, pop.description, detect.1,
                         detect.2, ddf.analyses, seed = 999)

tables$data["Region.Label"] <- rep("study.area", length(tables$data$object))


results <- boot.dht(tables, B=49, trunc=30)
results.df <- rbind(results$original, results$bootstrap)
kable(results.df)

```

### Abundance

True abundance of 20 animals.

``` {r 20 animals, echo=FALSE, warning=FALSE}
lines.med <- data.frame(x   = c(rep(seq(0, 1000, len=n_segs), 9)),
                        y   = c(seq(100, 100, len=n_segs),
                                seq(200, 200, len=n_segs),
                                seq(300, 300, len=n_segs),
                                seq(400, 400, len=n_segs),
                                seq(500, 500, len=n_segs),
                                seq(600, 600, len=n_segs),
                                seq(700, 700, len=n_segs),
                                seq(800, 800, len=n_segs),
                                seq(900, 900, len=n_segs)),
                        leg = c(rep("1", n_segs),
                                rep("2", n_segs),
                                rep("3", n_segs),
                                rep("4", n_segs),
                                rep("5", n_segs),
                                rep("6", n_segs),
                                rep("7", n_segs),
                                rep("8", n_segs),
                                rep("9", n_segs)))

unlink(shapefile.pathway, recursive = TRUE)
write.transects(lines.med, shapefile.pathway)

survey.design <- make.design(transect.type = "Line",
                             design.details = c("Parallel","Systematic"), 
                             region = region, plus.sampling = FALSE, 
                             path = shapefile.pathway)

pop.description <- make.population.description(N = 20, density.obj = pop.density,
                                               region = region, fixed.N = TRUE)

tables <- double.obs.sim(region, survey.design, pop.description, detect.1,
                         detect.2, ddf.analyses)

tables$data["Region.Label"] <- rep("study.area", length(tables$data$object))


results <- boot.dht(tables, B=49, trunc=30)
results.df <- rbind(results$original, results$bootstrap)
kable(results.df)

```

True abundance of 500 animals

``` {r 500 animals, echo=FALSE}
pop.description <- make.population.description(N = 500, density.obj = pop.density,
                                               region = region, fixed.N = TRUE)

tables <- double.obs.sim(region, survey.design, pop.description, detect.1,
                         detect.2, ddf.analyses, seed = 999)

tables$data["Region.Label"] <- rep("study.area", length(tables$data$object))


results <- boot.dht(tables, B=49, trunc=30)
results.df <- rbind(results$original, results$bootstrap)
kable(results.df)
```

True abundance of 2000 animals

``` {r 2000 animals, echo=FALSE}
pop.description <- make.population.description(N = 2000, density.obj = pop.density,
                                               region = region, fixed.N = TRUE)

tables <- double.obs.sim(region, survey.design, pop.description, detect.1,
                         detect.2, ddf.analyses, seed = 999)

tables$data["Region.Label"] <- rep("study.area", length(tables$data$object))


results <- boot.dht(tables, B=49, trunc=30)
results.df <- rbind(results$original, results$bootstrap)
kable(results.df)
```

### Density

We will consider three density surfaces describing the distribution of animals within the survey region. The plots are shown below.

``` {r density plots, echo=FALSE, fig.width=10, fig.height=4}
par(mfrow=c(1,3))

pop.density.centre <- make.density(region.obj = region, x.space = 10,
                                   y.space = 10, constant = 0.5)
pop.density.centre <- add.hotspot(pop.density.centre, centre = c(500, 500),
                                  sigma = 1000, amplitude = 0.2)


pop.density.west <- make.density(region.obj = region, x.space = 10,
                                 y.space = 10, constant = 0.5)
pop.density.west <- add.hotspot(pop.density.west, centre = c(200, 500),
                                sigma = 250, amplitude = 2)


pop.density.cluster <- make.density(region.obj = region, x.space = 10,
                                    y.space = 10, constant = 0.5)
pop.density.cluster <- add.hotspot(pop.density.cluster, centre = c(250, 250),
                                   sigma = 100, amplitude = 2)
pop.density.cluster <- add.hotspot(pop.density.cluster, centre = c(750, 750),
                                   sigma = 100, amplitude = 2)


plot(pop.density.centre)
plot(pop.density.west)
plot(pop.density.cluster)

par(mfrow=c(1,1))
```

Centre

``` {r centre density, echo=FALSE}
pop.description <- make.population.description(N = 500, density.obj = pop.density.centre,
                                               region = region, fixed.N = TRUE)

tables <- double.obs.sim(region, survey.design, pop.description, detect.1,
                         detect.2, ddf.analyses, seed = 999)

tables$data["Region.Label"] <- rep("study.area", length(tables$data$object))


results <- boot.dht(tables, B=49, trunc=30)
results.df <- rbind(results$original, results$bootstrap)
kable(results.df)
```

West

``` {r west density, echo=FALSE}
pop.description <- make.population.description(N = 500, density.obj = pop.density.west,
                                               region = region, fixed.N = TRUE)

tables <- double.obs.sim(region, survey.design, pop.description, detect.1,
                         detect.2, ddf.analyses)

tables$data["Region.Label"] <- rep("study.area", length(tables$data$object))


results <- boot.dht(tables, B=49, trunc=30)
results.df <- rbind(results$original, results$bootstrap)
kable(results.df)
```

Cluster

``` {r cluster density, echo=FALSE}
pop.description <- make.population.description(N = 500, density.obj = pop.density.cluster,
                                               region = region, fixed.N = TRUE)

tables <- double.obs.sim(region, survey.design, pop.description, detect.1,
                         detect.2, ddf.analyses, seed = 999)

tables$data["Region.Label"] <- rep("study.area", length(tables$data$object))


results <- boot.dht(tables, B=49, trunc=30)
results.df <- rbind(results$original, results$bootstrap)
kable(results.df)
```

[^1]: Adapted from the vignette by Laura Marshall:   
      http://www.creem.st-and.ac.uk/DS.M&A/simulation/DSsimstudy.html#directory-structure-for-files-in-this-project
      
[^2]: write.transects() function by David Miller:   
      https://github.com/dill/spatlaugh/blob/master/write_transects.R
      
[^3]: double.obs.sim() function:   
      https://github.com/m-murchie/dht-bootstrap/blob/master/R/double.obs.sim.R